---
title: "Visualizing Identity: A Study of Sexual Orientation and Religion Across England and Wales"
output: html_document
df_print: paged
toc: TRUE
code_folding: TRUE
date: 
---

# Visualizing Identity: A Study of Sexual Orientation and Religion Across England and Wales

## Data

The data for this project was downloaded from the Office for National Statistics website (link in references). The original file (which can be found in the GitHub repository's data file) includes the code book and metadata on sheets on the corresponding sheets.

The data was a product of the 2021 census – the sheets downloaded contained other demographic data that were collected alongside sexual orientation at the point of the census including: age, sex, employment status, smoking habits and general health.

## Research question

The decision to focus on religion and sexual orientation came after a recent discussion that ensued after sharing religious school experiences with friends and colleagues. The question being that, even if one found themselves of a sexual orientation that was deemed anathema by their church/religion, would they still identify with that religion? (which inevitably led to many other questions, outlined below under follow-up). Some churches and religious branches are now very progressive, but a religion, in its very nature, cannot be. Before scrolling through the ONS website I proposed that the data would represent this with a higher percentage of LGBTQIA+ groups identifying as not religious than the heterosexual population.

## Project script

#### Packages and importing the dataset

I developed a loop to automatically check the installation of necessary packages, ensuring the script can operate seamlessly on any system without manual adjustments. I first established the *packages* object, which includes the names of all required packages for the project script.

```{r echo=TRUE, message=FALSE, warning=FALSE}
packages <- c("here", "readxl", "tidyverse", "shiny", "vcd", "viridis", "scales", "plotly")
```

The *for* loop iterates over each package vector element, using the variable *pkg* to hold the current package name in each iteration.

```{r echo=TRUE, message=FALSE, warning=FALSE}
for (pkg in packages) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg)
    library(pkg, character.only = TRUE)
  }
}
```

Inside the loop, the *require()* function checks if the package named by *pkg* is installed and can be loaded. The *character.only = TRUE* argument specifies that *pkg* is a character string (the name of the package). While the *quietly = TRUE* argument suppresses warnings and other messages (to keep the console output clean as it's checking for multiple packages). If *require()* returns *FALSE* (the package is not available), *install.packages(pkg)* is called to install the package.

After installing necessary packages*, library()* is used to load all of them into the session. The *character.only = TRUE* argument again, indicates that *pkg* is a character string.

The following code then defines the path to the xlxs file using the *here() function.*

```{r echo=TRUE, message=FALSE, paged.print=TRUE}
excel_file <- here("data", "sexualorientationfurtherpersonalcharacteristicsenglandandwalescensus2021.xlsx")


sexualorientationreligion <- read_excel(excel_file, 
                                        sheet = "2a", 
                                        col_types = c("skip", "text", "text", "text", "text", "text", "numeric"), 
                                        skip = 4)
```


```{r interactive_table_1, echo=FALSE, message=FALSE, warning=FALSE}
library(DT)

datatable(sexualorientationreligion,
  options = list(pageLength = 5, autoWidth = TRUE),
  caption = "Raw Data")

    
```

After which sheet 2a of the file is read into the session skipping the top 4 rows and first column then defining which columns are text/numeric.

#### Preparing the data

After viewing the dataset to ensure that the correct columns/rows had been imported and all observations were there, I then began cleaning and filtering. The console was displaying the following warning.

``` r
Warning message:
Expecting numeric in G426 / R426C7: got '[c]'
```

After consulting the ONS metadata I discovered that '[c]' was in the column if the value was less than 10 so it was suppressed. To correct this I created a cleaned dataframe where all cells not including [c] were retained while those including [c] were excluded.

```{r echo=TRUE, message=FALSE, warning=FALSE}
sexualorientationreligion_clean <- sexualorientationreligion[!apply(sexualorientationreligion, 1, function(x) any(grepl("\\[c\\]", x))), ]

```

Following that I renamed the variables, see code below.

```{r paged.print=TRUE, echo=TRUE}
sexualorientationreligion_clean <- rename(sexualorientationreligion_clean,    
                                          c( "sexual_orientation" = "Sexual orientation", 
                                             "age" = "Age group [note 2]" , 
                                             "sex" = "Sex [note 1]", 
                                             "percentage" = "Percentage estimate of group \r\n[note 3] [note 4]"))

```

I then used the *dplyr* filtering function to create a new dataframe including only the data that was necessary for the final plot.

```{r}
filtered_data <- sexualorientationreligion_clean %>%
  filter(sexual_orientation != "All usual residents", 
         age == "All ages 16 years and over", 
         sex == "People")  
```

*filtered_data* is the new dataframe created to store the result of *dplyr'*s filtering operation. This new dataframe contains only the rows from sexualorientationreligion_clean that meet the specified conditions. The pipe operator passes the *sexualorientationreligion_clean* dataframe into the next function, *filter()* which specifies the following conditions:

-   *sexual_orientation != "All usual residents"* excludes the columns under *sexual_orientation* containing the listed value including only the breakdown of sexual orientations rather than the total of all.

-   *age == "All ages 16 years and over"* ensures that only rows where the age column exactly matches "All ages 16 years and over" are retained, excluding the columns breaking the data down by age group.

-   *sex == "People"* filters to include only rows where the sex column has the value "People", excluding any rows that might have other breakdowns by sex.

## Filtered data

```{r interactive-table, echo=FALSE, message=FALSE, warning=FALSE}
# Display the dataframe as an interactive table
library(DT)
datatable(filtered_data,
  options = list(pageLength = 5, autoWidth = TRUE),
  caption = "Final Dataframe")


```
## Plot

```{r echo=TRUE, fig.height=10, fig.width=10, paged.print=TRUE}

p <- ggplot(filtered_data, aes(x = sexual_orientation, y = percentage / 100, fill = Religion, text = paste("Percentage:", scales::percent(percentage / 100)))) +
  geom_bar(stat = "identity", position = "stack") +
  coord_flip() +
  labs(title = "Sexual Orientation and Religious Identity",
       x = "Sexual Orientation",
       y = "Percentage") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "Pastel1") +
  theme_light() +
  theme(legend.position = "bottom")

interactive_plot <- ggplotly(p, tooltip = "text")  

interactive_plot
  
```


#### Interpretation

## Future work and follow-up
